---
title: 'Reproducible Research: Peer Assessment 1'
output: html_document
keep_md: yes
---

## Software Environment
```{r software_env, echo=FALSE, warning=FALSE, message=TRUE}
sessionInfo()
```

## Loading and preprocessing the data

First of all, needed library are loaded.<br/>
Then, activies data are retrieved from activity.csv file.<br/>
And finally, the date on string format are converted in Date class.<br/>

```{r library_loading, echo=TRUE, warning=FALSE, message=FALSE}
library(dplyr)
library(lattice)

activities <- read.csv("activity.csv", stringsAsFactors = FALSE)
activities$date <- as.Date(activities$date)
```



## What is mean total number of steps taken per day?

The data previously obtained are grouped by date,<br/>
then, for each day, the sum of step of each interval is calculated.<br/>
The histogram of the total number of steps can now be drawed.

```{r hist_total_number_steps, echo=TRUE, warning=FALSE, message=FALSE, fig.align='center', tidy=FALSE}
totalNumberStepsPerDay <- activities %>% 
                          group_by(date)  %>% 
                          summarise(daily = sum(steps, na.rm = TRUE)) 
hist(totalNumberStepsPerDay$daily, breaks=20, 
     main = "Histogram of total number of steps taken per day", 
     xlab = "Total number of steps")
```

With this list of total number of steps for each day, the mean and median can now be easily retrieved.
```{r mean_total_number_steps, echo=TRUE, warning=FALSE, message=FALSE}
meanStep <- as.integer(mean(totalNumberStepsPerDay$daily, na.rm = TRUE))
medianStep <- as.integer(median(totalNumberStepsPerDay$daily, na.rm = TRUE))
```
The integer value of mean of total step is `r meanStep`.<br/>
The median of total step is `r medianStep`.


## What is the average daily activity pattern?

Again, The data previously obtained are used, but now grouped by interval,<br/>
then, for each interval, the mean of step calculated across all days values.<br/>
The time series plot  can now be drawed.

```{r plot_average_daily_activity, echo=TRUE, warning=FALSE, message=FALSE, fig.align='center', tidy=FALSE}
averageStepsPerInterval <- activities %>% 
                           group_by(interval) %>% 
                           summarise(averageInterval = mean(steps, na.rm = TRUE)) 
plot(as.numeric(averageStepsPerInterval$interval), 
                averageStepsPerInterval$averageInterval, 
                type="l", main = "Time series plot", 
                xlab = "5-minutes interval", ylab = "Average number of steps")
```
```{r max_average_daily_activity, echo=TRUE, warning=FALSE, message=FALSE}
maxInterval <- averageStepsPerInterval[which.max(averageStepsPerInterval$averageInterval), 
                                       "interval"][[1]]
``` 
The 5-minute interval, on average across all the days in the dataset, containing the maximum number of step is `r maxInterval`.<br/>

## Imputing missing values

First of all, calculate the sum of the NA values.
```{r NA_number, echo=TRUE, warning=FALSE, message=FALSE}
naNumber <- sum(is.na(activities))
```
The total number of missing values in the dataset is `r naNumber`.<br/>
As the average daily activity has been previously calculated, <br/>each NA value will be replaced by the mean of the corresponding interval.

```{r NA_replace, echo=TRUE, warning=FALSE, message=FALSE, tidy=FALSE}
emptyIndexes <- which(is.na(activities$steps))
completeActivities <- activities
completeActivities[emptyIndexes,]$steps <-
  merge(completeActivities[emptyIndexes,],averageStepsPerInterval)$averageInterval
```

Now, we have completed the dataset, we will build again the histogram :

```{r complete_hist_total_number_steps, echo=TRUE, warning=FALSE, message=FALSE, fig.align='center', tidy=FALSE}
totalNumberStepsPerDay <- completeActivities %>% 
                          group_by(date)  %>% 
                          summarise(daily = sum(steps, na.rm = TRUE)) 
hist(totalNumberStepsPerDay$daily, breaks=20,
     main = "Histogram of total number of steps taken per day", 
     xlab = "Total number of steps")
```

With this list of total number of steps for each day, the mean and median can be retrieved again.
```{r complete_mean_total_number_steps, echo=TRUE, warning=FALSE, message=FALSE}
meanStep <- as.integer(mean(totalNumberStepsPerDay$daily, na.rm = TRUE))
medianStep <- as.integer(median(totalNumberStepsPerDay$daily, na.rm = TRUE))
```
The integer value of mean of total step is `r meanStep`.<br/>
The median of total step is `r medianStep`.

So, with complete data, the mean and median of the total daily number of steps <br/>
 is higher than with original data with NA value.


## Are there differences in activity patterns between weekdays and weekends?

Firstly, we create a function that give a new factor  with two levels - "weekday" and "weekend" indicating whether a given date is a weekday or weekend day. <br/>
Secondly, we apply it at complete data and creating a new column for it.

```{r weekend_weekday, echo=TRUE, warning=FALSE, message=FALSE, tidy=FALSE}
getWeekPeriod <- function(d) {
    if(weekdays(d) %in% c("Saturday", "Sunday")) {
        return("weekend")
    }
    else {
        return("weekday")
    }
}

completeActivities <- cbind(completeActivities, 
                            weekPeriod = mapply(getWeekPeriod, completeActivities$date))
```

Finally, by grouping it by interval and weekPeriod, we split data to draw the plot for weekday and weekend. 

```{r final_plot_average_daily_activity, echo=TRUE, warning=FALSE, message=FALSE, fig.align='center'}
completeActivities <- completeActivities %>% 
                 group_by(interval, weekPeriod)  %>% 
                 summarise(averageInterval = mean(steps, na.rm = TRUE))


xyplot(averageInterval~interval | weekPeriod,
                    type="l", layout=c(1, 2), 
                    data = completeActivities, as.table = TRUE,
                    xlab = "5-minutes Interval", ylab = "Number of steps", 
                    main = "Time serie plot with weekend/weekday split")
```

So we can observe, that in weekend, the morning peak around 8h35 AM is not as important as it is during week days.